---
title: "Heading for Trouble? Concussions in College Soccer"
subtitle: "What 391 NCAA Concussions Reveal About Injury Causes and Recovery Timelines"
author: "Romy Bos"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-title: "On this page"
execute:
  echo: false
  warning: false
  message: false
---

## The Campus Game: What's Really Causing Concussions?

Soccer is one of the most popular sports on college campuses, but head injuries remain a serious hazard. When people imagine a sport-related concussion (SRC), they usually picture two players crashing together mid-air or someone violently striking the turf. 

A comprehensive study from the **NCAA-DoD CARE Consortium** (*Jo et al., 2024*, published in *The American Journal of Sports Medicine*) analyzed **391 soccer concussions** across collegiate programs to uncover how players actually get hurt—and how long it takes them to get back on the pitch.

::: {.callout-important appearance="simple"}
### Surprising Finding
Heading the ball isn't just routine soccer play—it is the **#1 single cause** of collegiate soccer concussions, accounting for nearly one in three injuries overall and **over 51%** of concussions sustained during team practice.
:::

## Part 1: How Do Concussions Happen?

The study classified every concussion into four main injury mechanisms:

1. **Head-to-Ball:** Direct contact with the ball (e.g., contested headers, high-speed deflections, or close-range shots).
2. **Head-to-Body:** Striking another player's shoulder, elbow, or torso during a challenge.
3. **Head-to-Head:** Direct skull-to-skull collisions with an opponent or teammate.
4. **Head-to-Ground / Equipment:** Hitting the turf, goalposts, or boundary barriers.

*Hover over any bar in the interactive chart below to see exact numbers:*

```{r}
library(tidyverse)
library(plotly)
library(scales)

mechanism_data <- tibble(
  mechanism = c(
    "Head-to-Ball",
    "Head-to-Body",
    "Head-to-Head",
    "Head-to-Ground / Equipment"
  ),
  count = c(128, 109, 85, 69),
  percentage = c(32.7, 27.9, 21.7, 17.6)
)

p1 <- mechanism_data |>
  ggplot(aes(
    x = percentage,
    y = fct_reorder(mechanism, percentage),
    text = paste0(
      "<b>", mechanism, "</b><br>",
      "Share of Total: ", percentage, "%<br>",
      "Concussions: ", count, " of 391"
    )
  )) +
  geom_col(aes(fill = percentage), show.legend = FALSE, width = 0.6) +
  scale_fill_gradient(low = "#5dade2", high = "#1b4f72") +
  scale_x_continuous(labels = scales::percent_format(scale = 1), limits = c(0, 40)) +
  labs(
    title = "Concussion Breakdown by Contact Mechanism",
    x = "Share of Total Concussions (%)",
    y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 14, margin = margin(b = 6)),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank()
  )

ggplotly(p1, tooltip = "text") |>
  config(displayModeBar = FALSE)
```

## Part 2: The Road to Recovery — How Long Are Athletes Sidelined?

How long does it take for a student-athlete to be fully cleared for **unrestricted return-to-play (URTP)**?

Athletes don't simply jump back into games right after feeling better. They must complete a supervised graduated return-to-play protocol, progressing from light aerobic exercise to sport-specific drills, non-contact practice, full practice, and finally game clearance.

Across all 391 cases, the average recovery time was **16.6 days**. However, recovery varied significantly depending on how the player was concussed:

*Hover over each bar below to compare average and median days out:*

```{r}
recovery_data <- tibble(
  mechanism = c(
    "Head-to-Ball",
    "Head-to-Head",
    "Head-to-Body",
    "Head-to-Ground / Equipment"
  ),
  mean_days = c(18.6, 16.9, 15.7, 14.0),
  median_days = c(13.9, 11.7, 12.4, 12.1)
)

p2 <- recovery_data |>
  ggplot(aes(
    x = mean_days,
    y = fct_reorder(mechanism, mean_days),
    text = paste0(
      "<b>", mechanism, "</b><br>",
      "Mean Recovery: ", mean_days, " days<br>",
      "Median Recovery: ", median_days, " days"
    )
  )) +
  geom_col(aes(fill = mean_days), show.legend = FALSE, width = 0.6) +
  scale_fill_gradient(low = "#48c774", high = "#e74c3c") +
  scale_x_continuous(limits = c(0, 24), breaks = seq(0, 24, by = 4)) +
  labs(
    title = "Average Days to Full Clearance (Unrestricted Return-to-Play)",
    x = "Average Days Sidelined",
    y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 14, margin = margin(b = 6)),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank()
  )

ggplotly(p2, tooltip = "text") |>
  config(displayModeBar = FALSE)
```

::: {.callout-note appearance="simple"}
### Why Did Ball Impacts Lead to Longer Recoveries?
Athletes who hit the ground or equipment recovered quickest (averaging **14.0 days**), while head-to-ball impacts kept athletes out the longest (averaging **18.6 days**). Researchers note that repetitive headers during training sessions can cause cumulative sub-concussive micro-trauma before a diagnosed concussion occurs, potentially slowing neurocognitive recovery.
:::

## Part 3: Practice vs. Game Day Reality

The danger profile changes drastically depending on whether athletes are in drills or playing a competitive match:

| Environment | Top Danger | Why It Happens |
|:---|:---|:---|
| **Practice** | **Head-to-Ball (51.1%)** | Repetitive heading drills, crosses, and close-quarter scrimmage shots. |
| **Competition** | **Player Collisions (66.8%)** | Full-throttle aerial challenges and contested loose balls (head-to-body: 36.0%, head-to-head: 30.8%). |

## Takeaways for College Athletes & Coaches

- **Smarter Training:** Limit high-repetition heading drills in training to curb cumulative head impacts.
- **Fair Play Enforcement:** Strict referee enforcement on reckless challenges in the air can reduce high-speed head-to-head collisions during match play.
- **Listen to Your Body:** Because ball-impact concussions can take nearly three full weeks to heal, athletes should never rush back through concussion protocol before fully clearing clinical benchmarks.

---
*Source: Jo, J., Boltz, A. J., Williams, K. L., et al. (2024). Mechanisms of Injury Leading to Concussions in Collegiate Soccer Players: A CARE Consortium Study. The American Journal of Sports Medicine, 52(6), 1585–1595.*
